Laurentia's continued motion into the Neoproterozoic: new paleomagnetic pole from the Jacobsville sandstone

### 2022-02-01 Manuscript/Manuscript.tex
format initial manuscript file in AGU template

\documentclass[11pt,letterpaper]{article}
\usepackage{xcolor}
\usepackage{textcomp,marvosym}
\usepackage{amsmath,amssymb}
\usepackage[left]{lineno}
\usepackage{changepage}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{sidecap}
\usepackage{pdfpages}
\usepackage{longtable}
\usepackage{url}

\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\doublespacing

\raggedright
\textwidth = 6.5 in
\textheight = 8.25 in
\oddsidemargin = 0.0 in
\evensidemargin = 0.0 in
\topmargin = 0.0 in
\headheight = 0.0 in
\headsep = 0.5 in
\parskip = 0.1 in
\parindent = 0.2 in

\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\lhead{}
\rhead{\thepage}

@@

\begin{flushleft}

{\Large \textbf{Laurentia's continued motion into the Neoproterozoic: new paleomagnetic pole from the Jacobsville sandstone}}
\\
\singlespacing
 Yiming Zhang\textsuperscript{1}, Nicholas Swanson-Hysell\textsuperscript{1}, Blake Hodgin\textsuperscript{1}, James Pierce\textsuperscript{1}, Anthony Fuentes\textsuperscript{1}\\
\bigskip
\textsuperscript{1} Department of Earth and Planetary Science, University of California, Berkeley, CA, USA\\

\smallskip
\end{flushleft}

%\linenumbers
%\pagestyle{empty}

@@

The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and the associated deposition of sediments in the interior of Laurentia while the continent traveled rapidly from high to low latitudes. Coeval with the cessation of rifting and decrease in the speed of plate motion is the ca. 1090-980 Ma Grenvillian Orogeny whose far-field stress is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation. A lack of paleomagnetic record from the interior of Laurentia between the end of the Midcontinent Rift and the cooling of rocks associated with the Grenvillian Orogeny resulted in a gap of record for the position and orientation of Laurentia for nearly 100 Myr. We develop a new paleomagnetic pole from from red beds of the Jacobsville Formation along 5 stratigraphic sections in Northern Michigan, US. High-resolution thermal demagnetization experiments together with successful fold tests resolve primary magnetic remanence which show shallow inclinations despite being  flattened. Inclination shallowing corrected paleomagnetic pole position places Laurentia at low latitudes south to the Keweenawan Track ca. 990 Ma, confirming continued plate motion of Laurentia across the equator in the Late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicate the Grenville Loop poles are younger than previously thought.

@@

Paleomagnetic data for the Jacobsville have led to a paleomagnetic pole position that appears to be a continuation of the Keweenawan Track from the Oronto Group poles (Roy and Robertson, 1978; Fig. 7; Table 4). As a result, its age has been interpolated to be latest Mesoproterozoic (e.g., 1020 Ma in Li et al., 2008) such that it was deposited prior to the ca. 1000 Ma Grenville loop poles. However, this age assignment has recently been questioned on the basis of laser-ablation–inductively cou- pled plasma–mass spectrometry (LA-ICP-MS) U-Pb detrital zircon dates from Jacobsville Sandstone samples. Malone et al. (2016) interpreted a maximum depositional age of 959 ± 19 Ma based on the weighted mean of the four youngest grains from 2050 LA-ICP-MS zircon dates. A similar age was proposed on the basis of the youngest LA-ICP-MS U-Pb dates in the study of Craddock et al. (2013). This young maximum interpreted age is intriguing because it suggests the unconformity above the Freda Sandstone spans at least 100 m.y. and was followed by continued sedi- mentation in the region long after rift activity. Nevertheless, the paleo- magnetic pole of the Jacobsville Sandstone (Roy and Robertson, 1978) appears to extend the Keweenawan Track from the Oronto Group poles in a manner that is consistent with deposition of the Jacobsville Sand- stone in the late Mesoproterozoic. Given the conflict with this inferred age, the large uncertainty of individual LA-ICP-MS dates, and the pos- sibility of outlier dates in such a large data set, it would be beneficial for the youngest Jacobsville zircons to be analyzed using CA-ID-TIMS to obtain more precise constraints.

The extensive paleomagnetic studies of the well-preserved late Mesoproterozoic to Neoproterozoic rocks of North America have provided the central record for global paleogeography reconstruction during the assembly and rearrangement of the postulated supercontinent Rodinia. Paleomagnetic studies of rocks from Laurentia have led to a series of paleomagnetic poles that form an apparent polar wander path (APWP) known as the ``Logan Loop" for the older high latitude poles at its apex that continues into the ``Keweenawan Track" of younger lower latitude poles that form a progression as the APWP heads toward the ``Grenville Loop" \citep{Swanson-Hysell2019a}. In particular, recent advances in pairing high-precision zircon U-Pb geochronology with high-quality paleomagnetic poles significantly improved the resolution of the Keweenawan Track and revealed Laurentia's rapid motion ($>$20 cm/yr) during the Midcontinent Rift magmatic activity from ca. 1110 Ma to ca. 1085 Ma \citep{Swanson-Hysell2019a}.

@@

It has been interpreted that the Jacobsville Sandstone is equivalent to the Bayfield Group in northern Wisconsin, the Fond Du Lac formation in Minnesota. Their shared lithofacies of conglomerate, sandstone, and siltstones as well as similar pattern of lithology transition from conglomerate dominated deposition to sandstone to occurrence of very fine grained siltstone facies  supports the interpretation that these sedimentary rocks formed during the period of thermal subsidence and inversion of the continental rifting after the major magmatism of the MCR.

@@

%\begin{figure}[!ht]
%\noindent\includegraphics[width=\textwidth]{}
%\centering
%\caption{\small{}}
%\label{fig:map}
%\end{figure}

@@

\subsection*{ACKNOWLEDGEMENTS}
%We thank James Pierce for his field assistance. 
%\footnotesize

\singlespacing
\clearpage
\bibliographystyle{gsabull}
\bibliography{YZ_ref}

>>>>

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers

\draftfalse

%% Enter journal name below.
%% Choose from this list of Journals:
%
% JGR: Atmospheres
% JGR: Biogeosciences
% JGR: Earth Surface
% JGR: Oceans
% JGR: Planets
% JGR: Solid Earth
% JGR: Space Physics
% Global Biogeochemical Cycles
% Geophysical Research Letters
% Paleoceanography and Paleoclimatology
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resources Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science
% Geohealth
%

\journalname{Geochemistry, Geophysics, Geosystems}

@@

\title{Laurentia's continued motion into the Neoproterozoic: new paleomagnetic pole from the Jacobsville sandstone}

\authors{Yiming Zhang\textsuperscript{1}, Nicholas Swanson-Hysell\textsuperscript{1}, Blake Hodgin\textsuperscript{1}, James Pierce\textsuperscript{1}, Anthony Fuentes\textsuperscript{1}}

\affiliation{1}{Department of Earth and Planetary Science, University of California, Berkeley, CA, USA}
% \affiliation{2}{Second Affiliation}
% \affiliation{3}{Third Affiliation}
% \affiliation{4}{Fourth Affiliation}

\correspondingauthor{Yiming Zhang}{yimingzhang@berkeley.edu}

\begin{keypoints}
\item A new paleomagnetic pole has been developed from the ca.990 Ma Jacobsville Formation 
\item Laurentia plate continued to move after 
\item The new pole from the Jacobsville Formation indicates the Grenville Loop could be younger than thought
\end{keypoints}

\begin{abstract}
[ enter your Abstract here ]
\end{abstract}

\section*{Plain Language Summary}
[ enter your Plain Language Summary here or delete this section]

@@

The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and the associated deposition of sediments in the interior of Laurentia while the continent traveled rapidly from high to low latitudes. Coeval with the cessation of rifting and decrease in the speed of plate motion is the ca. 1090-980 Ma Grenvillian Orogeny whose far-field stress is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation. A lack of paleomagnetic record from the interior of Laurentia between the end of the Midcontinent Rift and the cooling of rocks associated with the Grenvillian Orogeny resulted in a gap of record for the position and orientation of Laurentia for nearly 100 Myr. To fill this gap, we develop a new paleomagnetic pole from from red beds of the Jacobsville Formation along 5 stratigraphic sections in Northern Michigan, US. High-resolution thermal demagnetization experiments together with successful fold tests resolve primary magnetic remanence which show shallow inclinations despite being  flattened. Inclination shallowing corrected paleomagnetic pole position places Laurentia at low latitudes south of the Keweenawan Track ca. 990 Ma, confirming continued plate motion of Laurentia across the equator in the Late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the Grenville Loop poles are younger than previously thought.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{../Figures/Geologic_map.pdf}
\caption{caption}
\label{pngfiguresample}
\end{figure}

@@

Paleomagnetic data for the Jacobsville Formation have led to a paleomagnetic pole position that appears to be a continuation of the Keweenawan Track from the Oronto Group poles (Roy and Robertson, 1978; Fig. 7; Table 4). As a result, its age has been interpolated to be latest Mesoproterozoic (e.g., 1020 Ma in Li et al., 2008) such that it was deposited prior to the ca. 1000 Ma Grenville loop poles. However, this age assignment has recently been questioned on the basis of laser-ablation–inductively cou- pled plasma–mass spectrometry (LA-ICP-MS) U-Pb detrital zircon dates from Jacobsville Sandstone samples. Malone et al. (2016) interpreted a maximum depositional age of 959 ± 19 Ma based on the weighted mean of the four youngest grains from 2050 LA-ICP-MS zircon dates. A similar age was proposed on the basis of the youngest LA-ICP-MS U-Pb dates in the study of Craddock et al. (2013). This young maximum interpreted age is intriguing because it suggests the unconformity above the Freda Sandstone spans at least 100 m.y. and was followed by continued sedimentation in the region long after rift activity. Nevertheless, the paleo- magnetic pole of the Jacobsville Sandstone (Roy and Robertson, 1978) appears to extend the Keweenawan Track from the Oronto Group poles in a manner that may be more consistent with deposition of the Jacobsville Formation in the late Mesoproterozoic. Given the conflict with this inferred age, the large uncertainty of individual LA-ICP-MS dates, and the pos- sibility of outlier dates in such a large data set, it is beneficial to interpret the age of the youngest Jacobsville zircons dated CA-ID-TIMS to obtain more precise constraints. Using CA-ID-TIMS, a recent study by Hodgin et al (2022) redefines the maximum deposition age of the Jacobsville Formation as ca. 993 Ma.

The extensive paleomagnetic studies of the well-preserved late Mesoproterozoic to Neoproterozoic rocks of North America have provided the central record for global paleogeography reconstruction during the assembly and rearrangement of the postulated supercontinent Rodinia. Paleomagnetic studies of rocks from Laurentia have led to a series of paleomagnetic poles that form an apparent polar wander path (APWP) known as the "Logan Loop" for the older high latitude poles at its apex that continues into the ``Keweenawan Track" of younger lower latitude poles that form a progression as the APWP heads toward the ``Grenville Loop" \citep{Swanson-Hysell2019a}. In particular, recent advances in pairing high-precision zircon U-Pb geochronology with high-quality paleomagnetic poles significantly improved the resolution of the Keweenawan Track and revealed Laurentia's rapid motion ($>$20 cm/yr) during the Midcontinent Rift magmatic activity from ca. 1110 Ma to ca. 1085 Ma \citep{Swanson-Hysell2019a}.

@@

It has been interpreted that the Jacobsville Sandstone is equivalent to the Bayfield Group in northern Wisconsin, and the Fond Du Lac and Hinckley formations in Minnesota. Their shared lithofacies of basal conglomerate, cross-bedded fluvial fine to medium grained sandstone, and interbedded red micaceous siltstone could  support the interpretation that these sedimentary rocks formed during a common period of post-rift basin formation following rift-related magmatism of the MCR.

@@

@@

\acknowledgments
Enter acknowledgments, including your data availability statement, here.

\bibliography{YZ_ref}
